Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-09-08 latex/intro.tex
Done?

\section{Introduction}\label{section:intro}
The practice of experimental design entails specifying every detail of
running an experiment \cite{chaloner1995}. In its most basic form,
experimental design entails specifying measurements that consist an
experiment. For example, consider the process of searching for oil:
measurements require digging holes in the ground and a \emph{design}
simply indicates which holes should be dug
\cite{horesh2008borehole}. Since measurements are expensive, only a
limited number of them can be taken, and thus carefully choosing
measurement locations is crucial.

@@

Inverse problems are ubiquitous in science and engineering. For
example, in electrical impedance tomography (EIT) we seek to infer the
structure of the inside of a human body. Inference is conducted based
on observations of electrical current impedance measured at specific
electrode locations on the skin, leveraging known equations of
electric current flow \cite{horesh2010impedance}. Magnetic Resonance
Imaging (MRI) also entails solving an inverse problem. There,
radio-frequency pulses are sent through the human body in the presence
of a strong magnetic field. The response of one's body contents to
these radio-frequency pulses is measured, allowing a radiologist to
view the internals of a body in a noninvasive manner
\cite{horesh2008mri}. An inverse problem also arises in oil and gas
exploration, where acoustic waves are sent through "boreholes" ---
deep cylindrical wells drilled into the ground. Data of travel and
return times of the acoustic waves are recorded. Wave travel and
return time is influenced by the properties of the subsurface
materials, such as density, elasticity, and the presence of fluids or
voids. Combining travel time data with a geophysical models of
contents of the earth's crust facilitates reconstructing the structure
of the subsurface in a process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola2005}. In many inverse problems the
goal is to infer some numerical parameter. However, in the formulation
we consider in this study, as well as in the examples above, the goal
is to conduct inference over some \emph{function} over $\Omega$, where
\(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain of
interest.

@@

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Leibler (KL) divergence from posterior
to prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}. Utilizing KL divergence for quantifying differences
between distributions is standard practice and we will not motivate it
further here. For a discrete parameter $\param$ and data $\data$, the
KL divergence is defined as
\begin{equation}\label{eq:basic_KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted, so we average over it to define the D-optimality criterion
as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ].
\end{equation*}
We refer to a set of measurements that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

@@

\subsection{A toy model: the 1D heat equation}\label{subsec:toy}
For concreteness, let us now consider a toy inverse problem: inferring
the initial condition for partial differential equation known as the
\emph{heat equation} in one dimension (1D heat equation
henceforth). Readers less familiar with partial differential equations
can think of the following setup: before us there is a perfectly
insulated metal rod. At each point on the rod the temperature is
different and unknown, and we call this temperature distribution the
\emph{initial condition}. We wait for a short time $T$ and let the
heat dissipate a bit inside the rod. The heat equation determines the
heat distribution inside the rod at time $t=T$.

@@

Solving the 1D heat equation is straightforward. Consider an initial
heat distribution $u_0\in L^2([0,1])$ that satisfies the homogeneous
Dirichlet boundary condition eq.~\eqref{eq:heat2}. The initial
condition $u_0$ is a linear combination of sines $\ev_n(x) = \sin(\pi
n x)$, so $u_0 = \sum_{n\geq 1} a_n \ev_n$ for some $\{a_n\}_{n\geq
  1}$. Now, note that $\ev_n$ are in fact eigenvectors of the
Laplacian: $\Delta \ev_n = -\pi^2 n^2\ev_n$. It should not be hard to
believe that $u(\cdot, T) = \sum_{n\geq 1} a_n \exp(-\pi^2 n^2T )
\ev_n$ \cite{renardy2006PDE}. Thus, eigenvectors of the linear
operator that describes the time evolution of heat are $\ev_n(x) =
\sin(\pi n x)$ with corresponding eigenvalues
\begin{equation}\label{eq:decay}
  \exp(-\pi^2 n^2T ).
\end{equation}

In the \emph{inverse problem of the 1D heat equation}, our goal is to
infer the initial condition $u_0 = u(\cdot, 0)$ from observations of
the final state $u(\cdot, T)$. However, we are not able to measure
$u(\cdot, T)$ at every point in the domain $\Omega$. Rather, we take
some number of temperature measurements on the rod once our waiting
time $T$ has passed and try to infer $u(\cdot, 0)$ from these
measurements.

@@

In order to find a fully Bayesian formulation for our inverse problem
we first need to specify a prior on \emph{functions} defined over
$\Omega = [0,1]$. For the sake of simplicity we would like to utilize
a Gaussian prior. While specifying the prior mean is easy --- we take
the zero function on $\Omega$ --- specifying a prior covariance is
more involved. In analogy with the finite dimensional case, we seek a
covariance \emph{operator} that is positive definite and imposes
sufficient regularity on functions over $\Omega$. Thus, we set our
prior for the initial condition $u_0 \sim \mathcal{N}(0,
(-\Delta)^{-1})$, where $\Delta$ again defined with homogeneous
Dirichlet boundary condition. The intuition here is that
$(-\Delta)^{-1}$ is smoothing and prior realizations are generated, in
analogy to the finite dimensional case, via
$(-\Delta)^{-1/2}\mathcal{W}$, where $\mathcal{W}$ is white noise
\cite{rue2011}. Our choice of prior covariance operator ensures the
posterior is well-defined and prior realizations are well-behaved, see
Theorem 3.1 and Lemma 6.25 in \cite{Stuart10} for details. The second
ingredient of a fully Bayesian formulation of the inverse problem is
likelihood. We simply assume centered iid Gaussian measurement error
model, giving rise to a the widely applicable Gaussian likelihood
function.

@@

%% Replication 1
Clusterization should not be confused with replication. Replication
requires that the experimentalist executes multiple trials under
circumstances that are \emph{nominally identical} \cite[Section
  1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
and even necessary aspect of optimal experimental design
\cite{fisher1949design, morris2011, schafer2001replication}. Similarly
to a design implementing replication, a clustered design reduces the
signal-to-noise ratio of the repeated measurements
\cite{telford2007brief}. The difference is that a clustered design
takes repeated measurements at the expense of other quantities not
measured at all.

%% Replication 2
For example, \cite{fisher1949design}, suggested repeating his famous
milk and tea experiment in order "to be able to demonstrate the
predominance of correct classifications in spite of occasional
errors". Unfortunately, in the experiments we consider, replication is
impossible. For example, in the MRI problem, we cannot generate an
individual nominally identical to the one we wish to scan.

%% Replication 3
To further illustrate the difference between replication and
clusterization, consider an experiment measuring the effect of
rainfall on grass growth \cite{fay2000rainfall}. The experiment
involved four rainfall manipulation "treatments" (i.e.~simulating
different timing and quantity of rainfall), each replicated three
times over a different plot of land. Indeed, it seems reasonable for
researchers to replicate the phenomenon they are trying to study. A
clustered design in such an experiment would imply the researchers
should take a repeated measurement \emph{on the same plot}, at the
expense of measuring grass growth in other plots. To conclude our
short discussion on replication vs.~clusterization: these are
fundamentally different concepts and even though replication is quite
intuitive, it is inapplicable to the inverse problems we consider in
this manuscript.

@@

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization mainly depends
  on how quickly the eigenvalues of the prior covariance in
  observation space decay. This decay depends mostly on how ill-posed
  the problem at hand is and does not depend much on the prior. See
  Section \ref{section:vanishing} --- particularly, Theorem
  \ref{thm:char} and the discussion following it. We also show results
  of numerical experiments, where simulations of our model give rise
  to D-optimal designs that exhibit clusterization with high
  probability in Section \ref{subsec:lemma_sims}. Thus, given the
  genericity of our model, we expect measurement clusterization to be
  a generic and ubiquitous phenomenon.

@@

  The connection to Carath\'eodory's Theorem also builds on the focus
  of measurement effort to the abovementioned select set of prior
  eigenvectors. This concentration allows us to move from an
  infinite-dimensional setting to finite dimensions, where the
  conditions of Carath\'eodory's Theorem hold. We conclude that a
  D-optimal design arises by weighting a small number of
  measurements. If the number of allowed measurements is larger than
  the small number dictated by Carath\'eodory's Theorem, we have
  excessive weight on some measurements, which can be interpreted as
  clusterization.

@@

%% Convergence
Given that D-optimal designs suffer from clusterization, it is natural
to ask whether D-optimal designs should be pursued at all. Other
authors have established convergence results for decaying measurement
error \cite{knapik2011}, space-filling measurements
\cite{teckentrup2020} and randomly sampled designs
\cite{nickl2023}. Unfortunately, D-optimal designs --- clustered
designs in particular --- do not fall into any of these categories.

Luckily, we can expect convergence from D-optimal designs (including
clustered designs). Convergence holds in the following sense: the
posterior uncertainty ellipsoid in $\hilo$ will contract to zero along
every eigenfunction of $\fwd \prcov \fwd^*$. See Section
\ref{subsub:convergence} for a proof based on Theorem \ref{thm:char}
--- the main theorem of this manuscript.

\subsubsection{Other Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the space of \emph{observations}, see Section
\ref{section:vanishing} for a precise statement. This is in accordance
with previous work by \cite{koval2020}, who showed that A-optimal
designs are best constructed in the space of observations.

>>>>

\section{Introduction}\label{section:intro}
Experimental design is a systematic approach to specifying every
possible aspect of an experiment \cite{chaloner1995}. In this study,
we focus on experiments that involve measuring quantities arising from
naturally occurring processes. For us, therefore, experimental design
entails determining which specific measurements to record from a given
process. In a typical experiment the number of measurements that can
be recorded is limited by time, cost and other constraints. Therefore
selecting the right set of measurements to record is crucial as it
directly impacts the utility of the data collected, and consequently,
the validity of the conclusions drawn.

For example, consider the process of searching for oil: measurements
require digging deep holes in the ground, known as "boreholes". A
\emph{design} simply indicates which boreholes should be dug
\cite{horesh2008borehole}. Since digging boreholes is expensive, only
a limited number of them can be dug, and thus carefully choosing their
locations is crucial.

@@

Inverse problems are ubiquitous in science and engineering. For
example, in electrical impedance tomography (EIT) we seek to infer the
structure of the inside of a human body. Inference is conducted based
on observations of electrical current impedance measured at specific
electrode locations on the skin, leveraging known equations of
electric current flow \cite{horesh2010impedance}. Magnetic Resonance
Imaging (MRI) also entails solving an inverse problem. There,
radio-frequency pulses are sent through the human body in the presence
of a strong magnetic field. The response of one's body contents to
these radio-frequency pulses is measured, allowing a radiologist to
view the internals of a body in a noninvasive manner
\cite{horesh2008mri}. An inverse problem also arises in oil and gas
exploration, where acoustic waves are sent through "boreholes" ---
deep cylindrical wells drilled into the ground. Data of travel and
return times of the acoustic waves are recorded. Wave travel and
return time is influenced by the properties of the subsurface
materials, such as density, elasticity, and the presence of fluids or
voids. Combining travel time data with a geophysical model of the
contents of the earth's crust facilitates reconstructing the structure
of the subsurface in a process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola2005}. In many inverse problems the
goal is to infer some numerical parameter. However, in the formulation
we consider in this study, as well as in the examples above, the goal
is to conduct inference over some \emph{function} over $\Omega$, where
\(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain of
interest.

@@

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Leibler (KL) divergence from posterior
to prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}.
%%  Utilizing KL divergence for quantifying differences between
%% distributions is standard practice and we will not motivate it
%% further here.
Recall that for a discrete parameter $\param$ and data $\data$, the KL
divergence is defined as
\begin{equation}\label{eq:basic_KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted. Hence, we average over $\data$ to define the D-optimality
criterion as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ].
\end{equation*}
We refer to a set of measurements that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

@@

\subsection{A toy model: the 1D heat equation}\label{subsec:toy}
For concreteness, let us now consider a toy inverse problem: inferring
the initial condition for a partial differential equation known as the
\emph{heat equation} in one dimension (1D heat equation
henceforth). Readers less familiar with partial differential equations
can think of the following setup: before us there is a metal rod,
perfectly insulated except for its tips. At each point on the rod the
temperature is different and unknown, and we call this temperature
distribution the \emph{initial condition}. We wait for a short time
$T$ and let the heat dissipate a bit inside the rod. The heat equation
determines the heat distribution inside the rod at time $t=T$.

@@

Solving the 1D heat equation is straightforward. Consider an initial
heat distribution $u_0\in L^2([0,1])$ that satisfies the homogeneous
Dirichlet boundary condition eq.~\eqref{eq:heat2}. This initial
condition $u_0$ is a linear combination of sines $\ev_n(x) = \sin(\pi
n x)$, so $u_0 = \sum_{n\geq 1} a_n \ev_n$ for some $\{a_n\}_{n\geq
  1}$. Now, note that $\ev_n$ are in fact eigenvectors of the
Laplacian: $\Delta \ev_n = -\pi^2 n^2\ev_n$. It should not be hard to
believe that $u(\cdot, T) = \sum_{n\geq 1} a_n \exp(-\pi^2 n^2T )
\ev_n$ \cite{renardy2006PDE}. Thus, eigenvectors of the linear
operator that describes the time evolution of heat are $\ev_n(x) =
\sin(\pi n x)$ with corresponding eigenvalues
\begin{equation}\label{eq:decay}
  \exp(-\pi^2 n^2T ).
\end{equation}

In the \emph{inverse problem of the 1D heat equation}, our goal is to
infer the initial condition $u_0 = u(\cdot, 0)$ from noisy
observations of the final state $u(\cdot, T)$. However, we are not
able to measure $u(\cdot, T)$ at every point in the domain
$\Omega$. Rather, we take some number of temperature measurements on
the rod once our waiting time $T$ has passed and try to infer
$u(\cdot, 0)$ from these measurements.

@@

In order to find a fully Bayesian formulation for our inverse problem
we first need to specify a prior on \emph{functions} defined over
$\Omega = [0,1]$. For the sake of simplicity we would like to utilize
a Gaussian prior. While specifying the prior mean is easy --- we take
the zero function on $\Omega$ --- specifying a prior covariance is
more involved. In analogy with the finite dimensional case, we seek a
covariance \emph{operator} that is positive definite and imposes
sufficient regularity on functions over $\Omega$. Thus, we set our
prior for the initial condition $u_0 \sim \mathcal{N}(0,
(-\Delta)^{-1})$, where $\Delta$ is again defined with a homogeneous
Dirichlet boundary condition. The intuition here is that since
$\Delta$ is a differential operator, it is "roughing" (the opposite of
smoothing) and so $(-\Delta)^{-1}$ is smoothing. Prior realizations
are generated, in analogy to the finite dimensional case, by smoothing
white noise: $(-\Delta)^{-1/2}\mathcal{W}$, where $\mathcal{W}$ is
white noise \cite{rue2011}. Our choice of prior covariance operator
ensures the posterior is well-defined and prior realizations are
well-behaved, see Theorem 3.1 and Lemma 6.25 in \cite{Stuart10} for
details. The second ingredient of a fully Bayesian formulation of the
inverse problem is likelihood. We simply assume centered iid Gaussian
measurement error model, giving rise to a the widely applicable
Gaussian likelihood function.

Now that we have a fully Bayesian formulation of our toy inverse
problem, we would like to find a D-optimal design. Defining D-optimal
designs over function spaces is somewhat mathematically
involved. Thus, to keep our discussion here focused, we will explore
these technical details later in Section
\ref{subsec:D_optimal_design}. For now, it suffices to know that
D-optimal designs can be defined for our problem.

@@

%% Replication 1
Clusterization should not be confused with replication. Replication
requires that the experimentalist executes multiple trials under
circumstances that are \emph{nominally identical} \cite[Section
  1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
and even necessary aspect of optimal experimental design
\cite{fisher1949design, morris2011, schafer2001replication}.
For example, \cite{fisher1949design}, suggested repeating his famous
milk and tea experiment in order "to be able to demonstrate the
predominance of correct classifications in spite of occasional
errors". Unfortunately, in the experiments we consider, replication is
impossible. For example, in the MRI problem, we cannot generate an
individual nominally identical to the one we wish to scan.

Similarly to a design implementing replication, a clustered design
reduces the signal-to-noise ratio of the repeated measurements
\cite{telford2007brief}. The difference is that a clustered design
takes repeated measurements at the expense of other quantities not
measured at all. For example, consider an experiment measuring the
effect of rainfall on grass growth \cite{fay2000rainfall}. The
experiment involved four rainfall manipulation "treatments"
(i.e.~simulating different timing and quantity of rainfall), each
replicated three times over different plots of land. Indeed, it seems
reasonable for researchers to replicate the phenomenon they are trying
to study. A clustered design in such an experiment would imply the
researchers should take a repeated measurement \emph{on the same
plot}, at the expense of measuring grass growth in other plots!

To conclude our short discussion on replication vs.~clusterization:
these are fundamentally different concepts and even though replication
is quite intuitive, it is inapplicable to the inverse problems we
consider in this manuscript.

@@

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization mainly depends
  on how quickly the eigenvalues of the prior covariance in
  observation space decay.
  %% This decay depends mostly on how ill-posed the problem at hand is
  %% and does not depend much on the prior
  See Section \ref{section:vanishing} --- particularly, Theorem
  \ref{thm:char} and the discussion following it. Furhtermore, in
  Section \ref{subsec:lemma_sims} we show results of numerical
  experiments, where simulations of our model give rise to D-optimal
  designs that exhibit clusterization with high probability. Thus,
  given the genericity of our model, we expect measurement
  clusterization to be a generic and ubiquitous phenomenon.

@@

  The connection to Carath\'eodory's Theorem also builds on the focus
  of measurement effort to the abovementioned select set of prior
  eigenvectors. This allows us to move from an infinite-dimensional
  setting to finite dimensions, where the conditions of
  Carath\'eodory's Theorem hold. We conclude that a D-optimal design
  arises by weighting a small number of measurements. If the number of
  allowed measurements is larger than the small number dictated by
  Carath\'eodory's Theorem, we have excessive weight on some
  measurements, which can be interpreted as clusterization.

@@

%% Convergence
Clusterization in D-optimal designs raises the question of whether
D-optimal designs should be pursued at all. While other authors have
established convergence results for decaying measurement error
\cite{knapik2011}, space-filling measurements \cite{teckentrup2020}
and randomly sampled designs \cite{nickl2023}, their results do not
hold for D-optimal designs. %% Unfortunately, D-optimal designs ---
%% clustered designs in particular --- do not fall into any of these
%% categories.

Luckily, we can expect convergence from D-optimal designs (including
clustered designs). Convergence holds in the following sense: the
posterior uncertainty ellipsoid will contract to zero along each
one of its eigenvectors. See Section \ref{subsub:convergence} for a
precise statement and a proof based on Theorem \ref{thm:char} --- the
main theorem of this manuscript.

\subsubsection{Other Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the space of \emph{observations}
%% , see Section \ref{section:vanishing} for a precise statement.
This is in accordance with previous work by \cite{koval2020}, who
showed that A-optimal designs are best constructed in the space of
observations.
